---
title: "TitanicFirstTry"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r load libraries and data}
library(tidyverse)
library(stringr)
library(ggplot2)
library(GGally)

train <- read.csv("train.csv")
test <-  read.csv("test.csv")
train <- as_tibble(train)
test <- as_tibble(test)

summary(train)
```

Age appears to have NA's

Let's add a feature for title

```{r add Title}
train <- mutate(train, Title = 
                  gsub("^.*, (.*?)\\..*$", "\\1",as.character(Name)))
test <- mutate(test, Title = 
                 gsub("^.*, (.*?)\\..*$", "\\1",as.character(Name)))
train <- select(train, -c(PassengerId,Cabin,Ticket))
test <- select(test, -c(PassengerId,Cabin,Ticket))
train[train$Embarked =="",]$Embarked <- "S" #assign the mode
```

We'll fix the ages later
In the meantime, let's see how certain variables affect survival

```{r first ggpairs}
ggpairs(data = select(train,-Name))
```

Too many levels in Title.

```{r}
summary(with(train, aov(Survived ~ Title)))
```

Title is significant, clearly

```{r plot Title}
ggplot(data = train, aes(x = Title, y = Survived)) + geom_bar(
    stat = "summary", fun.y = mean) + coord_flip() +
    geom_text(aes(label = ..count.., y= ..prop..), 
    stat= "count", hjust = 1, col = "blue")
```

Some of these could certainly be combined - try Mrs and Miss, Mr and Master

```{r proportion tests}

Mrs <- filter(train, Title %in% c("Mrs"))
Miss <- filter(train, Title %in% c("Miss"))
counts <- c(dim(filter(Mrs))[1],dim(filter(Miss))[1])
lived <- c(dim(filter(Mrs,Survived == 1))[1],dim(filter(Miss,Survived == 1))[1])

Mr <- filter(train, Title %in% c("Mr"))
Master <- filter(train, Title %in% c("Master"))
counts2 <- c(dim(filter(Mr))[1],dim(filter(Master))[1])
lived2 <- c(dim(filter(Mr,Survived == 1))[1],dim(filter(Master,Survived == 1))[1])

prop.test(lived,counts)$p.value
prop.test(lived2,counts2)$p.value

```

I'll assume a strict(er) standard here, 90% significance.

At 90% significance we would reject both null hypotheses that these means are the same, so we should not merge them.

None of the rest of the "Title" values have a sufficient number of values to run statistical tests. Thus, title will merge the following:

Gentlelady <- the Countess, Ms, Mrs, Mme, Mlle, Lady
Miss
Gentleman <- Sir, Master, Major, Dr, Col
Mr <- Rev, Mr, Jonkheer, Don, Capt

In the test set, we'll go with

Gentlelady <- Ms, Mrs
Miss <- Dona, Miss
Gentleman <- Master, Dr, Col
Mr <- Mr, Rev

```{r grouping Titles}
train <- train %>% 
  mutate(Title = replace(Title, Title %in% c("the Countess", "Ms", "Mrs", "Mme", "Mlle", "Lady"),"Gentlelady")) %>% 
  mutate(Title = replace(Title, Title %in% c("Sir", "Master", "Major", "Dr", "Col"),"Gentleman")) %>% 
  mutate(Title = replace(Title, Title %in% c("Rev", "Mr", "Jonkheer", "Don", "Capt"),"Mr"))
  
test <- test %>% 
  mutate(Title = replace(Title, Title %in% c("Ms", "Mrs"),"Gentlelady")) %>% 
  mutate(Title = replace(Title, Title %in% c("Dona", "Miss"),"Miss")) %>% 
  mutate(Title = replace(Title, Title %in% c("Master", "Dr", "Col"),"Gentleman")) %>% 
  mutate(Title = replace(Title, Title %in% c("Mr", "Rev"),"Mr"))

ggplot(data = train, aes(x = Title, y = Survived)) + geom_bar(
    stat = "summary", fun.y = mean) + coord_flip() +
    geom_text(aes(label = ..count.., y= ..prop..), 
    stat= "count", hjust = 1, col = "blue")

```

That's more like it. Let's check Miss vs. Gentlelady just to be sure...

```{r Miss vs. Gentlelady}
Gentlelady <- filter(train, Title %in% c("Gentlelady"))
Miss <- filter(train, Title %in% c("Miss"))
counts <- c(dim(filter(Gentlelady))[1],dim(filter(Miss))[1])
lived <- c(dim(filter(Gentlelady,Survived == 1))[1],dim(filter(Miss,Survived == 1))[1])

prop.test(lived,counts)$p.value
```

P-value is lower than before, indicating that the probabilities of survival for these groups is even more divergent.

```{r ggpairs again}
train <- select(train, - Name)
ggpairs(data = train)

```

